Modelling vegetation around the world's first farms

### 2025-08-15 analysis/appendix.qmd
Finalise resubmission

---
title: "Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka"
subtitle: "Fitted model summaries and predicted palaeodistributions"
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This appendix describes ecological niche models of 69<!-- TODO keep updated -->
  plant species relevant to the subsistence of Late Epipalaeolithic (15–11.7 ka) 
  and Neolithic (11.7–8.2 ka) societies in West Asia. It includes summaries of 
  the training data and performance metrics of the fitted models together with 
  predicted palaeodistributions of each species for three simulated 
  climatologies over the Pleistocene/Holocene boundary (the Bølling–Allerød, 
  14.7–12.9 ka; Younger Dryas, 12.9–11.7 ka; and Early Holocene, 11.7–8.3 ka) as
  well as current conditions.
  For a full description of the methodology, references, and discussion of the 
  results, please see the main text. The data and R code used to produce the
  models, as well as the full predicted palaeodistributions in raster format,
  are achived with Zenodo at <https://doi.org/10.5281/zenodo.14629984>.
format:
  html:
    toc: true
    # embed-resources: true
execute:
  echo: false
---

>>>>

---
title: "Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka"
subtitle: "Fitted model summaries and predicted palaeodistributions"
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This appendix describes ecological niche models of 65<!-- TODO keep updated -->
  plant species relevant to the subsistence of Late Epipalaeolithic (15–11.7 ka) 
  and Neolithic (11.7–8.2 ka) societies in West Asia. It includes summaries of 
  the training data and performance metrics of the fitted models together with 
  predicted palaeodistributions of each species for three simulated 
  climatologies over the Pleistocene/Holocene boundary (the Bølling–Allerød, 
  14.7–12.9 ka; Younger Dryas, 12.9–11.7 ka; and Early Holocene, 11.7–8.3 ka) as
  well as current conditions.
  For a full description of the methodology, references, and discussion of the 
  results, please see the main text. The data and R code used to produce the
  models, as well as the full predicted palaeodistributions in raster format,
  are achived with Zenodo at <https://doi.org/10.5281/zenodo.14629984>.
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: false
---

### 2025-08-15 analysis/lay_summary.qmd
Finalise resubmission

We reconstructed the past ranges of 82 plant species that were used by prehistoric societies in West Asia around 10,000 years ago, including the ancestors of the first crops.
Our results suggest that under the cold, arid conditions of that time these were much less widespread than they are now, and that they became even scarcer just before the emergence of agriculture.
The Mediterranean coast of the Levant and, to our surprise, Cyprus and Western Anatolia appear to have been refugia for many economically important plants.
Our data introduces an as-yet unexplored new line of evidence on early plant domestication and has implications for how archaeologists interpret the botanical remains they recover from sites, as for example the recovery of a plant from outside its 'native range' is often taken as evidence that it was being managed or cultivated.
It also gives insights into the broader ecosystems from which the world's first agricultural systems emerged.

>>>>

We used machine learning to reconstruct the past ranges of 65 plant species that were used by prehistoric societies in West Asia around 10,000 years ago, including the ancestors of the first crops.
Our results suggest that under the cold, arid conditions of that time these were much less widespread than they are now, and that they became even scarcer just before the emergence of agriculture.
The Mediterranean coast of the Levant and, to our surprise, Cyprus and Western Anatolia appear to have been refugia for many economically important plants.
Our data introduces an as-yet unexplored new line of evidence on early plant domestication and has implications for how archaeologists interpret the botanical remains they recover from sites, as for example the recovery of a plant from outside its 'native range' is often taken as evidence that it was being managed or cultivated.
It also gives insights into the broader ecosystems from which the world's first agricultural systems emerged.
